### 2021-07-08 rad-icfp-2021/intro.tex
fixing boundary issues
reorganizing sections 5 and 6
reworked the introduction a bit

In this paper, we define a compilation scheme for a purely functional, provably correct, and efficient reverse-mode AD. 
To do so, we go start from a simple and general representation, where we can relatively easily prove correctness of reverse-mode.
Then, we define the UNF representation inspired by SSA~\cite{cytron1989efficient} and ANF, that leads to more efficient implementations.

@@

    \begin{center}
            \begin{tabular}{l}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !let y$_1$,y$_2$,y$_3$,y$_4$,y$_5$ = 0,0,0,0,1 in!\\
                !let y$_1'$! != y$_1$+w$_1$*y$_5$ in!\\
                !let y$_4'$! != y$_4$+x$_1$*y$_5$ in!\\
                !(y$_1'$+x$_2$*y$_4'$!!, y$_2$+x$_1$*y$_4'$!!, y$_3$)!
            \end{tabular}
    \end{center}   
For the expert viewer, this is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
This term can be further optimized via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !(w$_1$+x$_2$*x$_1$, x$_1$*x$_1$, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

\begin{itemize}[leftmargin=*]
\item We present our work with a simple yet expressive array-based language (with constructs such as !map2! and !reduce!) in Section~\ref{sec:simplediff}. 
We show how to directly compute an efficient reverse-mode AD for the expressions of this program (top of Figure~\ref{fig:outline}).  
Furthermore, we show how to extend our work to a richer language in Section~\ref{sec:generalization}.
\item One of the key insights behind efficient reverse-mode AD is to only consider unary operators. 
Inspired by this insight and following IRs such as SSA and ANF, we introduce a novel IR, UNF (Section~\ref{sec:unf})).
We introduce an alternative and easier to follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
\item We proof the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining the denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
\end{itemize}

Next, we recall rudiments of automatic differentiation, forward and reverse-mode differentiation

>>>>

In this paper, we define a purely functional (without references or control mechanisms such as state monads), denotationally correct, and provably efficient reverse-mode AD. 
To do so, we define the Unary Normal Form (UNF) representation inspired by PROPs \cite{hackney2015category} and compilation to categories \cite{elliott2017compiling}.
We can easily define and prove correctness of reverse-mode on this representation. 
The whole reverse-mode transformation is obtained by compiling the language to this intermediate representation, applying the simpler reverse-mode transformation, and compiling again to the original language.
After standard optimizations, the output program looks like SSA~\cite{cytron1989efficient} or ANF~\cite{sabry1993reasoning}, which leads to more efficient implementations.

@@

    \begin{center}
            \begin{tabular}{l}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !let y$_1$,y$_2$,y$_3$,y$_4$,y$_5$ = 0,0,0,0,1 in!\\
                !let y$_1'$! != y$_1$+w$_1$*y$_5$ in!\\
                !let y$_4'$! != y$_4$+x$_1$*y$_5$ in!\\
                !(y$_1'$+x$_2$*y$_4'$!!, y$_2$+x$_1$*y$_4'$!!, y$_3$)!
            \end{tabular}
    \end{center}   
This is equivalent to the SSA form \cite{cytron1989efficient} of what the imperative reverse-mode differentiation of our initial term would be.
This term can be further optimized via constant propagation and algebraic simplifications to give
        \begin{center}
            \begin{tabular}{{c}}
                !let w$_1$ = x$_1$ * x$_2$ in!\\ 
                !let w$_2$ = w$_1$ * x$_1$ in!\\
                !(w$_1$+x$_2$*x$_1$, x$_1$*x$_1$, 0)!
            \end{tabular}
        \end{center}
    \end{example}

@@

\begin{itemize}[leftmargin=*]
\item We present our work with a simple yet expressive array-based language (with constructs such as !map2! and !reduce!) in Section~\ref{sec:simplediff}. 
We show how to directly compute an efficient reverse-mode AD for the expressions of this program (top of Figure~\ref{fig:outline}).  
Furthermore, we show how to extend our work to a richer language in Section~\ref{sec:generalization}.
\item One of the key insights behind efficient reverse-mode AD is to only consider unary operators. 
Inspired by this insight and following IRs such as SSA and ANF, we introduce a novel IR, UNF (Section~\ref{sec:unf})).
We introduce an alternative and easier to follow compilation pipeline for efficient reverse-mode AD (bottom of Figure~\ref{fig:outline}).
\item We prove the complexity guarantees for the reverse ADed programs. Furthermore, we show a list of optimizations that can further improve the constant factors (Section~\ref{sec:complexity}).
\item We proof the correctness of our transformations (top/bottom parts of Figure~\ref{fig:outline}) by defining the denotational semantics of our languages using multicategories and concategories (Section~\ref{sec:correctness}).
\end{itemize}

Next, we recall rudiments of automatic differentiation, forward and reverse-mode differentiation.
